A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-10 tex/BiSig.tex
[ tex ] Revise S5 further
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

\usepackage{subfiles}
\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage[capitalise,noabbrev]{cleveref}

@@

\AtEndPreamble{%
\newtheorem*{claim}{Claim}
\theoremstyle{acmdefinition}
%\newtheorem*{claimproof}{Proof of Claim}
\newenvironment{claimproof}[1][\proofname]{\par
  \renewcommand{\qedsymbol}{\footnotesize$\blacksquare$}%
    \begin{proof}[#1]%
}{%
  \end{proof}%
}
\newtheorem{remark}[theorem]{Remark}
}

@@

\begin{abstract}
  \LT{Still not coherent enough}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  \LT{mode decorator?}
  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
\end{abstract}

>>>>

\usepackage{subfiles}
\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
%\usepackage{tikz}
%\usetikzlibrary{matrix,arrows}
\usepackage{mathpartir}

\usepackage{xspace}
\usepackage[capitalise,noabbrev]{cleveref}

@@

\AtEndPreamble{%
\newtheorem*{claim}{Claim}
\theoremstyle{acmdefinition}
%\newtheorem*{claimproof}{Proof of Claim}
\newenvironment{claimproof}{\par
  \renewcommand{\qedsymbol}{\footnotesize$\blacksquare$}%
  \begin{proof}[{Proof of Claim}]%
}{%
  \end{proof}%
}
\newtheorem{remark}[theorem]{Remark}
}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but only on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a first theory of bidirectional type synthesis.
The whole theory works as a verified type-synthesiser generator for syntax-directed simple type systems:
From a signature that specifies a bidirectionally typed language, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
Within the theory, we formally define soundness, completeness, and mode-correctness, which are sufficient conditions for deriving a correct type synthesiser.
We also propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations and streamlines the theory.
The entire theory is formulated constructively, and has been formalised in the \Agda proof assistant.

%\LT{Still not coherent enough}
%Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
%In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
%First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%which is analogous to a grammar in parsing.
%Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
%Moreover, we introduce `mode decoration' to deal with missing type annotations, an aspect often overlooked in the literature.
%The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
%By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
\end{abstract}

### 2023-07-10 tex/sec1-intro.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
More formally, judgements in a bidirectional type system are extended with two \emph{modes}:
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{enumerate*}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does---in fact, introducing long-distance unification constraints would undermine the essence of locality in bidirectional type synthesis.

@@

Although the basic idea is well known, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we present a first theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.

\subsection{Related Work}\label{sec:related-work}

@@

Within the theory, we formulate essential conditions including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
Practically, by running a mode preprocessor, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.

In summary, we contribute a theory of bidirectional type synthesis that is
\begin{enumerate}
  \item \emph{simple yet general}, working for any syntax-directed bidirectional simple type system that can be specified by a bidirectional binding signature;
  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode preprocessing over annotatability.
\end{enumerate}
Our theory was first developed with \Agda and then translated to the mathematical vernacular.

This paper is structured in logical order as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode preprocessing and related properties in \cref{sec:pre-synthesis}.
The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
In \cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
We conclude in \cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.

>>>>

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is assigned one of two \emph{modes}:
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{enumerate*}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does---in fact, introducing long-distance unification constraints would undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we present a first theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.

\subsection{Background and Related Work}\label{sec:related-work}

@@

Within the theory, we formulate essential conditions including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode decoration}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
Practically, by running a mode decorator, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
By combining bidirectional type synthesis, soundness, completeness, and mode decoration, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.

In summary, we contribute a theory of bidirectional type synthesis that is
\begin{enumerate}
  \item \emph{simple yet general}, working for any syntax-directed bidirectional simple type system that can be specified by a bidirectional binding signature;
  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode decoration over annotatability.
\end{enumerate}
Our theory was first developed with \Agda and then translated to the mathematical vernacular.

This paper is structured in logical order as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
In \cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
We conclude in \cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.

### 2023-07-10 tex/sec2-outline.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

\begin{definition}
\label{def:STLC-mode-preprocessor}
A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
\end{definition}

One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

\begin{theorem}\label{thm:implementation}
A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.
\end{theorem}

The construction is straightforward:
Run the mode preprocessor on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

@@

We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their generic definitions, which we formulate next.

%\begin{figure}
%  \small
%  \begin{tikzpicture}
%  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
%  {
%    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
%    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
%  };
%  \path[->, line width=0.5pt]
%      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
%      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
%      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
%      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
%  \path[dashed,->,line width=0.5pt]
%      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
%  \end{tikzpicture}
%  \caption{An implementation of type synthesis via bidirectional typing}
%\end{figure}

>>>>

\begin{definition} \label{def:STLC-mode-decorator}
A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
\end{definition}

One (less important) benefit of mode decoration is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

\begin{theorem}\label{thm:implementation}
A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
\end{theorem}

The construction is straightforward:
Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

@@

We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their generic definitions, which we formulate next.

%\begin{figure}
%  \small
%  \begin{tikzpicture}
%  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
%  {
%    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
%    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
%  };
%  \path[->, line width=0.5pt]
%      (m-1-1) edge node [above] {mode decoration} (m-1-2)
%      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
%      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
%      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
%  \path[dashed,->,line width=0.5pt]
%      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
%  \end{tikzpicture}
%  \caption{An implementation of type synthesis via bidirectional typing}
%\end{figure}

### 2023-07-10 tex/sec3-defs.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

For simple types, the only datum needed for specifying a type construct is its number of arguments:
\begin{definition} \label{def:simple-signature}
  \begin{figure}
%    \begin{minipage}[b]{.6\textwidth}
      \centering
      \small
      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
      \begin{mathpar}
        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
      \end{mathpar}
      \caption{Simple types}
      \label{fig:simple-type}
%    \end{minipage}
%    \begin{minipage}[b]{.35\textwidth}
%      \centering
%      \small
%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
%      \begin{mathpar}
%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
%      \end{mathpar}
%      \caption{Contexts}
%    \label{fig:simple-context}
%    \end{minipage}
  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimate sets in the sense of homotopy type theory.
  }
  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$,
  \end{enumerate}
  as shown in \Cref{fig:simple-type}.
  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
\end{definition}

@@

A simply typed language consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding:
\begin{definition}\label{def:binding-arity}
  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
  For brevity, a binding arity is denoted by $\bargs \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
\end{definition}

For example, consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is
\[
  {[A]B} \to {(A \bto B)}.
\]
This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$, contains:
\begin{enumerate*}
  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a \emph{context extended} by variable $x$ of type $A$;
  \item the type $A\bto B$ for itself.
\end{enumerate*}
In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$, saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.

>>>>

For simple types, the only datum needed for specifying a type construct is its number of arguments:
\begin{definition} \label{def:simple-signature}
  \begin{figure}
%    \begin{minipage}[b]{.6\textwidth}
      \centering
      \small
      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
      \begin{mathpar}
        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
      \end{mathpar}
      \caption{Simple types}
      \label{fig:simple-type}
%    \end{minipage}
%    \begin{minipage}[b]{.35\textwidth}
%      \centering
%      \small
%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
%      \begin{mathpar}
%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
%      \end{mathpar}
%      \caption{Contexts}
%    \label{fig:simple-context}
%    \end{minipage}
  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimate sets in the sense of homotopy type theory.
  }
  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$,
  \end{enumerate}
  as shown in \Cref{fig:simple-type}.
  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$.%
\todo{Name-type pairs?}
\end{definition}

@@

A simply typed language consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding:
\begin{definition}\label{def:binding-arity}
  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
  For brevity, a binding arity is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if $\Delta_i$ is empty.
\end{definition}

For example, consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is
\[
  {[A]B} \to {(A \bto B)}.
\]
This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
\begin{enumerate*}
  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a \emph{context extended} by variable $x$ of type $A$;
  \item the type $A\bto B$ for itself.
\end{enumerate*}
In a similar vein, the arity of the $\Rule{App}$ rule is denoted as
\[
  {(A \bto B), A} \to {B}
\]
saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
These arguments have types $A \bto B$ and $A$ respectively and empty extension contexts.
